---
title: "00_data cleaning"
author: "Brenna Kelly"
date: "2024-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# background
...

# set-up

```{r}

library(sf)
library(tmap)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(tidycensus)

```


## data

### aqs monitors
This code:
- 

```{r}

mons <- read.csv("/Users/brenna/Downloads/aqs_monitors-3.csv")

mons <- mons |>
  filter(Parameter.Code %in% c(44201, # o3
                               42401, # so2
                               42101, # co
                               42602, # no2
                               88101, # pm2.5
                               81102, 86101, # pm10
                               14129, 85129, 12128)) # lead

mons$lat_lon <- paste0(mons$Latitude, mons$Longitude)

mons$criteria <- case_when(mons$Parameter.Code %in% c(14129, 85129, 12128) ~ "pb",
                           mons$Parameter.Code == 42101 ~ "co",
                           mons$Parameter.Code %in% c(88101, 81102, 86101) ~ "pm",
                           mons$Parameter.Code == 42602 ~ "no2",
                           mons$Parameter.Code == 44201 ~ "o3",
                           mons$Parameter.Code == 42401 ~ "so2")

mons <- mons |>
  filter(!is.na(criteria))
mons$criteria <- as.factor(mons$criteria)

mons$last_year <- str_sub(mons$Last.Sample.Date, start = 0, end = 4)
mons$Last.Sample.Date <- ymd(mons$Last.Sample.Date)

# if there was a measurement within the past 5 years; sites are evaluated every 5 years
mons <- mons |>
  filter(last_year %in% c("", "2023", "2022", "2021", "2020", "2019"))

# imputing lat/long based on site number
mons$State.Code <- str_pad(mons$State.Code, pad = "0", width = 2, side = "left")
mons$County.Code <- str_pad(mons$County.Code, pad = "0", width = 3, side = "left")
mons$Site.Number <- str_pad(mons$Site.Number, pad = "0", width = 4, side = "left")

mons$site_code <- paste0(mons$State.Code, mons$County.Code, mons$Site.Number)

missing_sites <- mons[which(is.na(mons$Latitude)), "site_code"]
missing_sites <- mons[which(mons$site_code %in% missing_sites), ]

# the sites with missing lat/lon have addresses
  # google maps to the rescue
missing_sites[, c("Address", "State.Name", "site_code")]

# making a dataframe for merge
missing_site_locs <- data.frame(Latitude_m = c(41.759142851119755, 28.847295692743412, 27.997397528059278, 44.301764986058686,
                                             46.33560595983247, 43.636043773953546, 43.61948993912684, 43.579846480926776,
                                             43.60106825076201, 43.631329594263555, 43.61519809470876, 43.25579488107274,
                                             45.95800312479897, 37.27107085242156, 43.636043773953546, 43.636043773953546,
                                             43.636043773953546, 43.636043773953546),
                                Longitude_m = c(-72.6812059865072, -82.47762724105901, -82.68835723156158, -69.76804768078843,
                                              -86.84501913332537, -83.83589061574071, -84.19734680224849, -84.24246288690706,
                                              -84.20426022923483, -84.21565144483327, -84.23534194236433, -86.2416138858149,
                                              -86.2464834021623, -79.94489398712126, -83.83589061574071, -83.83589061574071,
                                              -83.83589061574071, -83.83589061574071),
                                site_code = c("090030019", "120170001", "121032002", "230110015",
                                              "260030901", "260170901", "261110901", "261110904",
                                              "261110905", "261110907", "261110910", "261210905",
                                              "261530901", "511610010", "260170902", "260170903", 
                                              "260170904", "260170905"))


mons <- merge(mons, missing_site_locs, by = c("site_code"), all.x = TRUE)

mons$Latitude <- ifelse(is.na(mons$Latitude),
                        mons$Latitude_m,
                        mons$Latitude)
mons$Longitude <- ifelse(is.na(mons$Longitude),
                         mons$Longitude_m,
                         mons$Longitude)

# AEA will be used for visualization, but distance-preserving projections like Mercator are needed for buffer calculations
aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +ellps=GRS80 +datum=NAD83"
mons <- st_as_sf(mons, coords = c("Longitude", "Latitude"),
                crs = 4326, agr = "constant")## |>
  #st_transform(3857) # mercator
mons <- st_transform(mons, st_crs(aea))

tm_shape(mons) +
  tm_dots(col = "criteria")

table(mons$Measurement.Scale.Definition,
      mons$Measurement.Scale)

# measurement scale is missing and not attainable for 1241 (16%) of observations
#   we'll impute with the median (neighborhood scale, 500M-4KM)

```


spatial

```{r}

# aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +ellps=GRS80 +datum=NAD83"
# 
# mons <- st_as_sf(mons, coords = c("Longitude", "Latitude"), 
#                  crs = 4326, agr = "constant")
# mons <- st_transform(mons, st_crs(aea))
# 
# tm_shape(mons) +
#   tm_dots(col = "criteria")
# 

```


### population characteristics

```{r}

acs_vars <- load_variables(2021, "acs5", cache = TRUE)


ST <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", 
        "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY",
        "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
        "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
        "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
        "VT", "VA", "WA", "WV", "WI", "WY") # for now, exclude U.S. territories

acs_dat <- list()

for (i in 1:length(ST)) {
    acs_df <- get_acs(geography = "block group",
                  state = ST[i],
                  variables = c("B03002_001",  # total pop, race denom
                                "B03002_012",  # hisp
                                "B03002_003",  # white
                                "B03002_004",  # black
                                "B03002_005",  # aian
                                "B03002_006",  # asian
                                "B03002_007",  # nhpi
                                "B03002_008",  # other
                                "B03002_009", # tom
                                "C17002_001",  # poverty denom 
                                "C17002_002", "C17002_003",  # below 1.00 
                                "B15003_001",  # ed denom 
                                "B15003_017", "B15003_018",  # hs, eq 
                                "B15003_019", "B15003_020", "B15003_021", # some college 
                                "B15003_022", "B15003_023", "B15003_024", "B15003_025"), # bach or higher 
                  geometry = TRUE,
                  year = 2021)
    acs_df <- acs_df |>
      group_by(GEOID) |>
      pivot_wider(values_from = "estimate", names_from = "variable") |>
      summarize(across(everything(), ~ first(na.omit(.)))) |>
      mutate(pov_below = C17002_002 + C17002_003) |>
      mutate(hs_eq = B15003_017 + B15003_018) |>
      mutate(some_coll = B15003_019 + B15003_020 + B15003_021) |>
      mutate(coll = B15003_022 + B15003_023 + B15003_024 + B15003_025) |>
      rename(total = B03002_001, hisp = B03002_012, white = B03002_003, black = B03002_004, 
             aian = B03002_005, asian = B03002_006, nhpi = B03002_007, other = B03002_008, 
             tom = B03002_009, pov_denom = C17002_001, 
             ed_denom = B15003_001) |>
      select(!starts_with("B1")) |>
      select(!starts_with("C1"))
    
    acs_dat[[i]] <- acs_df
}

acs_data <- do.call(rbind, acs_dat)

empty_units <- acs_data[which(st_is_empty(acs_data) == TRUE), ]
summary(empty_units) # these units have no population

acs_data <- acs_data |>
  mutate(hisp_p = hisp / total) |>
  mutate(white_p = white / total) |>
  mutate(black_p = black / total) |>
  mutate(aian_p = aian / total) |>
  mutate(asian_p = asian / total) |>
  mutate(nhpi_p = nhpi / total) |>
  mutate(other_p = other / total) |>
  mutate(tom_p = tom / total) |>
  mutate(pov_p = pov_below / pov_denom) |>
  mutate(hs_eq_p = hs_eq / ed_denom) |>
  mutate(some_coll_p = some_coll / ed_denom) |>
  mutate(coll_p = coll / ed_denom) |>
  mutate(no_hs_p = 1 - ((hs_eq + some_coll_p + some_coll_p) / ed_denom))

names(acs_data) <- tolower(names(acs_data))
  
write.csv(acs_data, "data/acs_data.csv", row.names = FALSE)

```

# population centers
We don't just want block groups; we want to know where the population is in that block group. We'll use population-weighted centroids later on to see if the measurement scale of EPA monitors intersects with these locations — in other words, is the majority of a population in an area likely captured within the monitor's measurement area.

```{r}

centroids <- read.csv("/Users/brenna/Documents/School/Research/aqs-inequities/aqs_inequities/CenPop2020_Mean_BG.csv")
names(centroids) <- tolower(names(centroids))
centroids <- st_as_sf(centroids, coords = c("longitude", "latitude"), 
                 crs = 4326, agr = "constant")## |>
  #st_transform(3857) # mercator

# creating block group fips for join
centroids$geoid <- paste0(str_pad(centroids$statefp, pad = "0", width = 2, side = "left"),
                          str_pad(centroids$countyfp, pad = "0", width = 3, side = "left"),
                          str_pad(centroids$tractce, pad = "0", width = 6, side = "left"),
                          centroids$blkgrpce)

```


## buffers
### creating buffers

```{r}

# stratify monitors by measurement scale
mons_micro <- mons[which(mons$Measurement.Scale == "MICROSCALE"), ]
mons_middle <- mons[which(mons$Measurement.Scale == "MIDDLE SCALE"), ]
mons_neighborhood <- mons[which(mons$Measurement.Scale %in% c("NEIGHBORHOOD", "")), ] # this is the imputation
mons_urban <- mons[which(mons$Measurement.Scale == "URBAN SCALE"), ]
mons_regional <- mons[which(mons$Measurement.Scale == "REGIONAL SCALE"), ]

 
# for sensitivity analysis in supplement
    # we imputed the most common scale for unknowns; but we'll test how much the results would change if all monitors were at the smallest and largest scales; i.e., we'll impute micro scale and regional scale
sa_mons_micro <- mons[which(mons$Measurement.Scale %in% c("MICROSCALE", "")), ]
sa_mons_regional <- mons[which(mons$Measurement.Scale %in% c("REGIONAL REGIONAL", "")), ]

# create the buffer for each measurement scale (in meters)

table(mons$Measurement.Scale.Definition, mons$Measurement.Scale)


micro_buffer_lower = st_buffer(mons_micro, 0.5)
micro_buffer_upper = st_buffer(mons_micro, 100)
middle_buffer_lower = st_buffer(mons_middle, 100)
middle_buffer_upper = st_buffer(mons_middle, 500)
neighborhood_buffer_lower = st_buffer(mons_neighborhood, 500)
neighborhood_buffer_upper = st_buffer(mons_neighborhood, 4000)
urban_buffer_lower = st_buffer(mons_urban, 4000)
urban_buffer_upper = st_buffer(mons_urban, 50000)
regional_buffer_lower = st_buffer(mons_regional,  50000)
regional_buffer_upper = st_buffer(mons_regional, 150000)

tm_shape(regional_buffer_upper) +
  tm_polygons(col = "criteria", alpha = 0.5, lwd = 0)

```

### stratifying buffers by criteria pollutant

```{r}

micro_buffer_lower_by_criteria <- split(micro_buffer_lower, micro_buffer_lower$criteria, 6)
micro_buffer_upper_by_criteria <- split(micro_buffer_upper, micro_buffer_upper$criteria, 6)
middle_buffer_lower_by_criteria <- split(middle_buffer_lower, middle_buffer_lower$criteria, 6)
middle_buffer_upper_by_criteria <- split(middle_buffer_upper, middle_buffer_upper$criteria, 6)
neighborhood_buffer_lower_by_criteria <- split(neighborhood_buffer_lower, neighborhood_buffer_lower$criteria, 6)
neighborhood_buffer_upper_by_criteria <- split(neighborhood_buffer_upper, neighborhood_buffer_upper$criteria, 6)
urban_buffer_lower_by_criteria <- split(urban_buffer_lower, urban_buffer_lower$criteria, 6)
urban_buffer_upper_by_criteria <- split(urban_buffer_upper, urban_buffer_upper$criteria, 6)
regional_buffer_lower_by_criteria <- split(regional_buffer_lower, regional_buffer_lower$criteria, 6)
regional_buffer_upper_by_criteria <- split(regional_buffer_upper, regional_buffer_upper$criteria, 6)

tm_shape(neighborhood_buffer_upper_by_criteria$pm) +
  tm_polygons(col = "criteria", lwd = 0)

```


### intersections by criteria pollutant
Here we're checking which population-weighted block group centroids lie inside the buffer (measurement scale of a monitor) based on the minimum and maximum scales given. At a later point we may want to use the median for simplicity, but also report the min/max? That may or may not simplify it.

We're also filtering out block groups which do not lie inside the buffers (observations where there is no `criteria` pollutant listed). We only want the block groups which lie inside. Then we can create binary variables (e.g., `upper_pm`, `lower_pm`) for each block group to tell us whether the block group is measured or not. Keep in mind, we don't care what scale the block group is measured at — only whether it was measured or not. Further down we'll combine the FIPS codes by pollutant across measurement scales. For instance, any FIPS with PM measured at any scale would be `upper_pm == "TRUE"`.

We only really need the block group FIPS code, but we'll select a few more variables which might be helpful

```{r}

centroids <- st_transform(centroids, st_crs(mons))

table(mons$Measurement.Scale, mons$criteria)
tm_shape(us_states) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_dots(col = "black") +
  tm_shape(micro_buffer_upper_by_criteria$pm) +
  tm_polygons(col = "red", alpha = 0.5)

tm_shape(micro_buffer_upper_by_criteria$pm[12, ]) +
  tm_polygons(col = "red", alpha = 0.5, lwd = 10) +
  
  tm_shape(centroids) +
  tm_dots(col = "black")

# PM
measured_bgc_micro_pm_upper <- st_join(centroids, micro_buffer_upper_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_pm_lower <- st_join(centroids, micro_buffer_lower_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_pm_upper <- st_join(centroids, middle_buffer_upper_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_pm_lower <- st_join(centroids, middle_buffer_lower_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_pm_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_pm_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_pm_upper <- st_join(centroids, urban_buffer_upper_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_pm_lower <- st_join(centroids, urban_buffer_lower_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_pm_upper <- st_join(centroids, regional_buffer_upper_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_pm_lower <- st_join(centroids, regional_buffer_lower_by_criteria$pm, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)

# SO2
measured_bgc_micro_so2_upper <- st_join(centroids, micro_buffer_upper_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_so2_lower <- st_join(centroids, micro_buffer_lower_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_so2_upper <- st_join(centroids, middle_buffer_upper_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_so2_lower <- st_join(centroids, middle_buffer_lower_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_so2_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_so2_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_so2_upper <- st_join(centroids, urban_buffer_upper_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_so2_lower <- st_join(centroids, urban_buffer_lower_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_so2_upper <- st_join(centroids, regional_buffer_upper_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_so2_lower <- st_join(centroids, regional_buffer_lower_by_criteria$so2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)

# NO2

table(is.na(measured_bgc_micro_no2_upper$criteria))
measured_bgc_micro_no2_upper <- st_join(centroids, micro_buffer_upper_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_no2_lower <- st_join(centroids, micro_buffer_lower_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_no2_upper <- st_join(centroids, middle_buffer_upper_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_no2_lower <- st_join(centroids, middle_buffer_lower_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_no2_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_no2_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_no2_upper <- st_join(centroids, urban_buffer_upper_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_no2_lower <- st_join(centroids, urban_buffer_lower_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_no2_upper <- st_join(centroids, regional_buffer_upper_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_no2_lower <- st_join(centroids, regional_buffer_lower_by_criteria$no2, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)

# CO
measured_bgc_micro_co_upper <- st_join(centroids, micro_buffer_upper_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_co_lower <- st_join(centroids, micro_buffer_lower_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_co_upper <- st_join(centroids, middle_buffer_upper_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_co_lower <- st_join(centroids, middle_buffer_lower_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_co_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_co_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_co_upper <- st_join(centroids, urban_buffer_upper_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_co_lower <- st_join(centroids, urban_buffer_lower_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_co_upper <- st_join(centroids, regional_buffer_upper_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_co_lower <- st_join(centroids, regional_buffer_lower_by_criteria$co, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)

# O3
measured_bgc_micro_o3_upper <- st_join(centroids, micro_buffer_upper_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_o3_lower <- st_join(centroids, micro_buffer_lower_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_o3_upper <- st_join(centroids, middle_buffer_upper_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_o3_lower <- st_join(centroids, middle_buffer_lower_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_o3_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_o3_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_o3_upper <- st_join(centroids, urban_buffer_upper_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_o3_lower <- st_join(centroids, urban_buffer_lower_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_o3_upper <- st_join(centroids, regional_buffer_upper_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_regional_o3_lower <- st_join(centroids, regional_buffer_lower_by_criteria$o3, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)

# PB
measured_bgc_micro_pb_upper <- st_join(centroids, micro_buffer_upper_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_micro_pb_lower <- st_join(centroids, micro_buffer_lower_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_pb_upper <- st_join(centroids, middle_buffer_upper_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_middle_pb_lower <- st_join(centroids, middle_buffer_lower_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_pb_upper <- st_join(centroids, neighborhood_buffer_upper_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_neighborhood_pb_lower <- st_join(centroids, neighborhood_buffer_lower_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_pb_upper <- st_join(centroids, urban_buffer_upper_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
measured_bgc_urban_pb_lower <- st_join(centroids, urban_buffer_lower_by_criteria$pb, join = st_within) |>
  filter(!is.na(criteria)) |>
  select(geoid, Measurement.Scale, 
         Measurement.Scale.Definition, 
         criteria)
# measured_bgc_regional_pb <- st_join(centroids, regional_buffer_upper_by_criteria$pb, join = st_within)
# no regional-scale lead monitors

```

let's make sure that worked properly

```{r}

# plotting the buffers
tm_shape(regional_buffer_upper_by_criteria$o3) +
  tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0) +
  tm_shape(micro_buffer_upper_by_criteria$o3) +
  tm_polygons(col = "criteria") +
  tm_shape(middle_buffer_upper_by_criteria$o3) +
  tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0) +
  tm_shape(neighborhood_buffer_upper_by_criteria$o3) +
  tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0) +
  tm_shape(urban_buffer_upper_by_criteria$o3) +
  tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0)

# plotting the centroids that lie within the buffers
tm_shape(measured_bgc_regional_o3_lower) +
  tm_dots(, col = "criteria", alpha = 0.3) +
  tm_shape(measured_bgc_urban_o3_lower) +
  tm_dots(col = "criteria") +
  tm_shape(measured_bgc_neighborhood_o3_lower) +
  tm_dots(, col = "criteria", alpha = 0.3)
  #tm_shape(measured_bgc_middle_o3_lower) +
  #tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0) +
  #tm_shape(measured_bgc_micro_o3_lower) +
  #tm_polygons(, col = "criteria", alpha = 0.3, lwd = 0)
tm_shape(measured_bgc_regional_o3_upper) +
  tm_dots(, col = "criteria", alpha = 0.3) +
  tm_shape(measured_bgc_urban_o3_upper) +
  tm_dots(col = "criteria") +
  tm_shape(measured_bgc_neighborhood_o3_upper) +
  tm_dots(, col = "criteria", alpha = 0.3)


```

### combining monitored FIPS by pollutant and range (upper/lower) across buffers

```{r}


measured_pm_upper <- rbind(measured_bgc_micro_pm_upper, measured_bgc_middle_pm_upper, 
                           measured_bgc_neighborhood_pm_upper, measured_bgc_urban_pm_upper,
                           measured_bgc_regional_pm_upper) |>
  filter(!duplicated(geoid))
measured_pm_lower <- rbind(measured_bgc_micro_pm_lower, measured_bgc_middle_pm_lower,
                           measured_bgc_neighborhood_pm_lower, measured_bgc_urban_pm_lower,
                           measured_bgc_regional_pm_lower) |>
  filter(!duplicated(geoid))

measured_so2_upper <- rbind(measured_bgc_micro_so2_upper, measured_bgc_middle_so2_upper, 
                           measured_bgc_neighborhood_so2_upper, measured_bgc_urban_so2_upper,
                           measured_bgc_regional_so2_upper) |>
  filter(!duplicated(geoid))
measured_so2_lower <- rbind(measured_bgc_micro_so2_lower, measured_bgc_middle_so2_lower,
                           measured_bgc_neighborhood_so2_lower, measured_bgc_urban_so2_lower,
                           measured_bgc_regional_so2_lower) |>
  filter(!duplicated(geoid))

measured_no2_upper <- rbind(measured_bgc_micro_no2_upper, measured_bgc_middle_no2_upper, 
                           measured_bgc_neighborhood_no2_upper, measured_bgc_urban_no2_upper,
                           measured_bgc_regional_no2_upper) |>
  filter(!duplicated(geoid))
measured_no2_lower <- rbind(measured_bgc_micro_no2_lower, measured_bgc_middle_no2_lower,
                           measured_bgc_neighborhood_no2_lower, measured_bgc_urban_no2_lower,
                           measured_bgc_regional_no2_lower) |>
  filter(!duplicated(geoid))

measured_co_upper <- rbind(measured_bgc_micro_co_upper, measured_bgc_middle_co_upper, 
                           measured_bgc_neighborhood_co_upper, measured_bgc_urban_co_upper,
                           measured_bgc_regional_co_upper) |>
  filter(!duplicated(geoid))
measured_co_lower <- rbind(measured_bgc_micro_co_lower, measured_bgc_middle_co_lower,
                           measured_bgc_neighborhood_co_lower, measured_bgc_urban_co_lower,
                           measured_bgc_regional_co_lower) |>
  filter(!duplicated(geoid))

measured_o3_upper <- rbind(measured_bgc_micro_o3_upper, measured_bgc_middle_o3_upper, 
                           measured_bgc_neighborhood_o3_upper, measured_bgc_urban_o3_upper,
                           measured_bgc_regional_o3_upper) |>
  filter(!duplicated(geoid))
measured_o3_lower <- rbind(measured_bgc_micro_o3_lower, measured_bgc_middle_o3_lower,
                           measured_bgc_neighborhood_o3_lower, measured_bgc_urban_o3_lower,
                           measured_bgc_regional_o3_lower) |>
  filter(!duplicated(geoid))

measured_pb_upper <- rbind(measured_bgc_micro_pb_upper, measured_bgc_middle_pb_upper, 
                           measured_bgc_neighborhood_pb_upper, measured_bgc_urban_pb_upper) |>
  filter(!duplicated(geoid))
measured_pb_lower <- rbind(measured_bgc_micro_pb_lower, measured_bgc_middle_pb_lower,
                           measured_bgc_neighborhood_pb_lower, measured_bgc_urban_pb_lower) |>
  filter(!duplicated(geoid))


```


## merging with population characteristics
Variables created and stored here need to be unique in the `acs_dat` dataframe.

```{r}

names(acs_dat) <- tolower(names(acs_dat))

measured_pb_lower$pb_scale_l <- paste0(measured_pb_lower$criteria, 
                                       ", ", measured_pb_lower$Measurement.Scale,
                                       ", ", "lower")
measured_pb_lower <- measured_pb_lower[, c("geoid", "pb_scale_l")] |>
  st_drop_geometry()

measured_pb_upper$pb_scale_u <- paste0(measured_pb_upper$criteria, 
                                       ", ", measured_pb_upper$Measurement.Scale,
                                       ", ", "upper")
measured_pb_upper <- measured_pb_upper[, c("geoid", "pb_scale_u")] |>
  st_drop_geometry()

measured_pm_lower$pm_scale_l <- paste0(measured_pm_lower$criteria, 
                                       ", ", measured_pm_lower$Measurement.Scale,
                                       ", ", "lower")
measured_pm_lower <- measured_pm_lower[, c("geoid", "pm_scale_l")] |>
  st_drop_geometry()

measured_pm_upper$pm_scale_u <- paste0(measured_pm_upper$criteria, 
                                       ", ", measured_pm_upper$Measurement.Scale,
                                       ", ", "upper")
measured_pm_upper <- measured_pm_upper[, c("geoid", "pm_scale_u")] |>
  st_drop_geometry()

measured_no2_lower$no2_scale_l <- paste0(measured_no2_lower$criteria, 
                                         ", ", measured_no2_lower$Measurement.Scale,
                                         ", ", "lower")
measured_no2_lower <- measured_no2_lower[, c("geoid", "no2_scale_l")] |>
  st_drop_geometry()

measured_no2_upper$no2_scale_u <- paste0(measured_no2_upper$criteria, 
                                         ", ", measured_no2_upper$Measurement.Scale,
                                         ", ", "upper")
measured_no2_upper <- measured_no2_upper[, c("geoid", "no2_scale_u")] |>
  st_drop_geometry()

measured_so2_lower$so2_scale_l <- paste0(measured_so2_lower$criteria, 
                                         ", ", measured_so2_lower$Measurement.Scale,
                                         ", ", "lower")
measured_so2_lower <- measured_so2_lower[, c("geoid", "so2_scale_l")] |>
  st_drop_geometry()

measured_so2_upper$so2_scale_u <- paste0(measured_so2_upper$criteria, 
                                         ", ", measured_so2_upper$Measurement.Scale,
                                         ", ", "upper")
measured_so2_upper <- measured_so2_upper[, c("geoid", "so2_scale_u")] |>
  st_drop_geometry()

measured_co_lower$co_scale_l <- paste0(measured_co_lower$criteria, 
                                       ", ", measured_co_lower$Measurement.Scale,
                                       ", ", "lower")
measured_co_lower <- measured_co_lower[, c("geoid", "co_scale_l")] |>
  st_drop_geometry()

measured_co_upper$co_scale_u <- paste0(measured_co_upper$criteria, 
                                       ", ", measured_co_upper$Measurement.Scale,
                                       ", ", "upper")
measured_co_upper <- measured_co_upper[, c("geoid", "co_scale_u")] |>
  st_drop_geometry()

measured_o3_lower$o3_scale_l <- paste0(measured_o3_lower$criteria, 
                                       ", ", measured_o3_lower$Measurement.Scale,
                                       ", ", "lower")
measured_o3_lower <- measured_o3_lower[, c("geoid", "o3_scale_l")] |>
  st_drop_geometry()

measured_o3_upper$o3_scale_u <- paste0(measured_o3_upper$criteria, 
                                       ", ", measured_o3_upper$Measurement.Scale,
                                       ", ", "upper")
measured_o3_upper <- measured_o3_upper[, c("geoid", "o3_scale_u")] |>
  st_drop_geometry()


acs_merge <- merge(acs_data, measured_pb_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_pb_upper, by = "geoid", all.x = TRUE) |>
  merge(measured_no2_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_no2_upper, by = "geoid", all.x = TRUE) |>
  merge(measured_pm_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_pm_upper, by = "geoid", all.x = TRUE) |>
  merge(measured_so2_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_so2_upper, by = "geoid", all.x = TRUE) |>
  merge(measured_o3_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_o3_upper, by = "geoid", all.x = TRUE) |>
  merge(measured_co_lower, by = "geoid", all.x = TRUE) |>
  merge(measured_co_upper, by = "geoid", all.x = TRUE) |>
  mutate(pb_measured_u = ifelse(is.na(pb_scale_u), 0, 1)) |>
  mutate(pb_measured_l = ifelse(is.na(pb_scale_l), 0, 1)) |>
  mutate(pm_measured_u = ifelse(is.na(pm_scale_u), 0, 1)) |>
  mutate(pm_measured_l = ifelse(is.na(pm_scale_l), 0, 1)) |>
  mutate(no2_measured_u = ifelse(is.na(no2_scale_u), 0, 1)) |>
  mutate(no2_measured_l = ifelse(is.na(no2_scale_l), 0, 1)) |>
  mutate(so2_measured_u = ifelse(is.na(so2_scale_u), 0, 1)) |>
  mutate(so2_measured_l = ifelse(is.na(so2_scale_l), 0, 1)) |>
  mutate(o3_measured_u = ifelse(is.na(o3_scale_u), 0, 1)) |>
  mutate(o3_measured_l = ifelse(is.na(o3_scale_l), 0, 1)) |>
  mutate(co_measured_u = ifelse(is.na(co_scale_u), 0, 1)) |>
  mutate(co_measured_l = ifelse(is.na(co_scale_l), 0, 1))

summary(acs_merge)

```

## save data

```{r}

acs_wo_geom <- st_drop_geometry(acs_merge)
acs_geom <- acs_merge |>
  select(geoid, geometry)

write.csv(acs_wo_geom, "data/7020_data.csv", row.names = FALSE)
st_write(acs_geom, "data/7020_geom.shp", append = TRUE)

```




